---
title: "Finding the AI conversation on Bluesky"
subtitle: "COMP3020 Social Web Analytics: exploratory phase"
author: "Trung, Thu and Christan (add all three student IDs before submitting)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.5)
```

# Design

The topic is not fixed in advance. The collection is broad, the patterns come out of the
text, and the focus is chosen from what the patterns show.

- Phase 1: collect posts on AI and related terms, month by month, so the sample covers a
  period rather than a single moment.
- Phase 2: find patterns with word frequency, TF-IDF and clustering.
- Phase 3: name the themes the clusters reveal and label every post with one.
- Phase 4: track the themes over time, then choose the focus topic for the poster.

Only the methods needed for these phases are included. Network analysis and link analysis
follow once the topic is chosen.

# Setup

## Loading Packages

```{r packages}
# Functions for retrieving Bluesky posts.
library(atrrr)
```

## Authenticating with Bluesky

```{r auth}
# Authorize the API to load dataset
 bsky_app_pw <-"vc7d-2zda-wr7q-oyb3"
 auth(user = "trungphanm.bsky.social", password = bsky_app_pw)
```

## Checking the Search Arguments

```{r check-args}
# The monthly collection needs both since and until. Confirm your version has them.
args(search_skeet)
```

If `until` is not listed, remove it from the search call below and filter on the post
timestamp afterwards. The monthly windows will then be uneven, because `sort = "latest"`
returns the most recent posts overall rather than the most recent inside a window.

# Phase 1: Collecting Bluesky Posts

## Searching Month by Month

Each phrase is searched once per month. Bluesky search takes one plain phrase at a time, so
several phrases are used to reach the whole conversation rather than one corner of it.

The collection is switched off once the data is saved. Set `eval=TRUE` on this chunk and the
next one to collect again.

```{r collect, eval=FALSE}
# WARNING: Bluesky content is uncontrolled and may contain offensive language.
queries = c("AI", "artificial intelligence", "generative AI", "ChatGPT")

# Bluesky opened to the public in February 2024, so earlier windows return nothing.
months = seq(as.Date("2024-01-01"), as.Date("2026-10-01"), by = "month")
per_call = 500

all_posts = NULL

for (q in queries) {
  for (i in 1:(length(months) - 1)) {
    res = search_skeet(q,
                       sort = "latest",
                       since = as.character(months[i]),
                       until = as.character(months[i + 1]),
                       limit = per_call
                       )
    if (nrow(res) > 0) {
      res$query = q
      res$window = format(months[i], "%Y-%m")
      all_posts = rbind(all_posts, res)
    }
  }
}
```

## Filtering and Saving

```{r filter-save, eval=FALSE}
# A post can match more than one phrase.
all_posts = all_posts[!duplicated(all_posts$uri), ]

# Keep English posts that contain text.
is_en = sapply(all_posts$langs, function(x) isTRUE(any(unlist(x) == "en")))
has_text = nchar(trimws(all_posts$text)) > 0
all_posts = all_posts[is_en & has_text, ]

save(all_posts, file = "AI_bluesky.RData")
```

## Loading the Saved Data

```{r load-data}
load("AI_bluesky.RData")

nrow(all_posts)
names(all_posts)
```

## Checking the Collection

```{r check-collection}
# Posts per phrase and per window. Any cell at the limit was cut off.
table(all_posts$query)
table(all_posts$window)
```

# Phase 2: Preparing the Text

## Creating the Corpus

```{r corpus}
library(tm)
library(SnowballC)

posts.text = all_posts$text
posts.corpus = Corpus(VectorSource(posts.text))

# Convert to ASCII to reduce encoding inconsistencies.
posts.corpus = tm_map(posts.corpus, function(x) iconv(x, to = "ASCII", sub = " "))
```

## Pre-processing the Corpus

```{r preprocess}
# Remove links first so domain fragments do not survive as terms.
remove_urls = content_transformer(function(x) {
  gsub("https?://\\S+|www\\.\\S+", " ", x, perl = TRUE)
})
posts.corpus = tm_map(posts.corpus, remove_urls)

posts.corpus = tm_map(posts.corpus, removeNumbers)
posts.corpus = tm_map(posts.corpus, removePunctuation)
posts.corpus = tm_map(posts.corpus, stripWhitespace)
posts.corpus = tm_map(posts.corpus, tolower)

# Only the words present in every post by construction are removed. Product names
# such as chatgpt are kept, because which tools people name is part of the pattern.
custom_words = c("ai", "artificial", "intelligence", "amp")
posts.corpus = tm_map(posts.corpus, removeWords,
                      c(stopwords("english"), custom_words))

posts.corpus = tm_map(posts.corpus, stemDocument)
```

# Phase 2: Finding Patterns

## Creating the Document-Term Matrix

```{r dtm}
posts.dtm = DocumentTermMatrix(posts.corpus)
dim(posts.dtm)

# With this many posts the full matrix cannot be densified, so the sparsest terms
# are dropped first. 0.99 keeps terms appearing in at least one percent of posts.
posts.dtm = removeSparseTerms(posts.dtm, 0.99)
posts.matrix = as.matrix(posts.dtm)
dim(posts.matrix)
```

## Top Terms by Frequency

```{r top-terms}
w = colSums(posts.matrix)
o = order(w, decreasing = TRUE)[1:30]
colnames(posts.matrix)[o]
```

## Top Terms by TF-IDF

```{r tfidf}
# Transpose the reduced matrix rather than rebuilding it from the corpus.
posts.tdm = t(posts.dtm)

posts.wtdm = weightTfIdf(posts.tdm)
posts.weighted.tdm = as.matrix(posts.wtdm)

sums = rowSums(posts.weighted.tdm)
o = order(sums, decreasing = TRUE)[1:30]
names(sums)[o]
```

## Word Clouds

```{r wordcloud-raw, fig.height=5.5}
library(wordcloud)

freqs = rowSums(as.matrix(posts.tdm))

wordcloud(names(freqs), freqs,
          random.order = FALSE, min.freq = 20, max.words = 150,
          colors = brewer.pal(8, "Dark2"), scale = c(3, 0.5))
title("Unweighted term frequencies")
```

```{r wordcloud-weighted, fig.height=5.5}
freqsw = rowSums(posts.weighted.tdm)

wordcloud(names(freqsw), freqsw,
          random.order = FALSE, min.freq = 0, max.words = 150,
          colors = brewer.pal(8, "Dark2"), scale = c(2.5, 0.4))
title("TF-IDF weighted terms")
```

## Clustering a Sample of Posts

Cosine distance needs an N by N matrix and MDS decomposes it, so the cost grows with the
square of the post count. The frequency work above uses every post. The clustering uses a
random sample.

```{r cluster-sample}
set.seed(3020)

max_cluster_posts = 2000
keep = sample(ncol(posts.wtdm), min(max_cluster_posts, ncol(posts.wtdm)))

# Clustering functions expect documents in rows.
posts.wmatrix = t(as.matrix(posts.wtdm[, keep]))

# Cleaning leaves some posts with no terms; zero rows carry no distance information.
empties = which(rowSums(abs(posts.wmatrix)) == 0)
if (length(empties) > 0) posts.wmatrix = posts.wmatrix[-empties, ]

dim(posts.wmatrix)
```

```{r cosine-mds}
# Normalise each document vector to unit length, then use cosine distance.
norm.posts.matrix = posts.wmatrix / sqrt(rowSums(posts.wmatrix^2))
D = dist(norm.posts.matrix, method = "euclidean")^2 / 2

# k-means only supports Euclidean geometry, so MDS projects the cosine distances.
N = nrow(posts.wmatrix)
mds.posts.matrix = cmdscale(D, k = min(100, N - 1))
dim(mds.posts.matrix)
```

```{r elbow}
n = 12
SSW = rep(0, n)
for (a in 1:n) {
  K = kmeans(mds.posts.matrix, a, nstart = 10)
  SSW[a] = K$tot.withinss
}

plot(1:n, SSW, type = "b",
     xlab = "Number of clusters", ylab = "Within-cluster sum of squares")
```

```{r kmeans}
# Read the elbow from the plot above before fixing this value.
number.of.clusters = 5

K = kmeans(mds.posts.matrix, number.of.clusters, nstart = 20)
table(K$cluster)

mds2.posts.matrix = cmdscale(D, k = 2)
plot(mds2.posts.matrix, col = K$cluster, pch = 19, cex = 0.5,
     xlab = "MDS dimension 1", ylab = "MDS dimension 2",
     main = "Sampled posts clustered by cosine distance")
```

```{r cluster-terms}
# Average each term's weight across a cluster. The highest means describe it, and
# they become the starting dictionary for the themes.
top_per_cluster = 10
theme_terms = list()

for (cluster.number in 1:number.of.clusters) {
  clusterpostsId = which(K$cluster == cluster.number)
  clusterTermWeight = colMeans(posts.wmatrix[clusterpostsId, , drop = FALSE])
  o = order(clusterTermWeight, decreasing = TRUE)[1:top_per_cluster]
  stems = colnames(posts.wmatrix)[o]

  theme_terms[[paste0("cluster", cluster.number)]] = stems
  cat("Cluster", cluster.number, "(", length(clusterpostsId), "posts):", stems, "\n")
}
```

# Phase 3: Naming the Themes

The dictionary comes from the clusters above, so the themes are taken from the data rather
than assumed. Read the cluster terms, then give each cluster a name.

```{r theme-names}
# Rename once the cluster terms have been read, for example:
# names(theme_terms) = c("work", "art", "study", "tools", "policy")

# Stems shared by several clusters cannot separate themes, so drop them.
all_stems = unlist(theme_terms)
shared = names(which(table(all_stems) > 1))
theme_terms = lapply(theme_terms, function(x) setdiff(x, shared))

theme_terms
```

```{r themes}
# A post takes the first theme it matches, so the order of the list matters.
cleaned.text = sapply(posts.corpus, as.character)

theme = rep("other", length(cleaned.text))
for (nm in names(theme_terms)) {
  stems = theme_terms[[nm]]
  if (length(stems) == 0) next
  pattern = paste0("\\b(", paste(stems, collapse = "|"), ")")
  hit = grepl(pattern, cleaned.text)
  theme[hit & theme == "other"] = nm
}

table(theme)
```

# Phase 4: Tracking the Themes Over Time

## Posts per Month

```{r trend-volume}
# Column names differ between atrrr versions, so resolve the timestamp first.
stamp = NULL
if ("created_at" %in% names(all_posts)) stamp = all_posts$created_at
if (is.null(stamp) && "indexed_at" %in% names(all_posts)) stamp = all_posts$indexed_at
if (is.null(stamp) && "post_data" %in% names(all_posts)) {
  stamp = sapply(all_posts$post_data, function(x) {
    v = unlist(x)
    hit = grep("^[0-9]{4}-[0-9]{2}-[0-9]{2}T", v, value = TRUE)
    if (length(hit) > 0) hit[1] else NA_character_
  })
}

all_posts$month = if (is.null(stamp)) NA_character_ else
  format(as.Date(substr(stamp, 1, 10)), "%Y-%m")

# Fall back to the search window if no timestamp survived.
if (all(is.na(all_posts$month))) all_posts$month = all_posts$window

sum(is.na(all_posts$month))

months.seen = sort(unique(all_posts$month))
counts = table(factor(all_posts$month, levels = months.seen))

plot(seq_along(counts), as.numeric(counts), type = "b",
     xaxt = "n", xlab = "", ylab = "Posts collected",
     main = "Posts collected per month")
axis(1, at = seq_along(counts), labels = names(counts), las = 2, cex.axis = 0.7)
```

Volume here reflects the collection limit as much as the conversation, so any month sitting
at the cap cannot be read as a real total. The share plot below is the honest version.

## Theme Share per Month

```{r trend-share}
tab = table(factor(all_posts$month, levels = months.seen), theme)
share = prop.table(tab, 1)

matplot(share, type = "b", pch = 1, lty = 1:ncol(share), col = 1:ncol(share),
        xaxt = "n", xlab = "", ylab = "Share of posts",
        main = "Theme share of the AI conversation")
axis(1, at = seq_len(nrow(share)), labels = rownames(share), las = 2, cex.axis = 0.7)
legend("topleft", colnames(share), col = 1:ncol(share), lty = 1:ncol(share),
       pch = 1, cex = 0.7, bty = "n")

round(share, 3)
```

# What to Decide Next

- Choose one or two themes that hold enough posts and show movement across the months.
- Run a focused collection on those themes with narrower phrases, using the same monthly loop.
- Add the network work from Modules 7 and 8 on the chosen theme: seed account, follows,
  centrality and PageRank.

# Limitations

- Each search is capped per month and per phrase, so counts measure collection effort as much
  as conversation volume. Read shares, not totals.
- Within a capped month, `sort = "latest"` returns posts from the end of that month, so the
  sample inside a capped window is not random.
- Only English posts are kept, and Bluesky users are not a cross-section of the public.
- Stemming merges word forms, so themes and cluster terms read as fragments.
- Themes are built from the top terms of a 2,000 post sample, then applied to every post.
- Theme labelling is first match wins, so overlapping posts land in the earlier theme. Report
  how many posts fall into "other".
