---
title: "The AI Conversation on Bluesky"
subtitle: "COMP3020 Social Web Analytics -- Hypothesis Testing, Text Analysis and Clustering"
author: "Trung, Thu and Christan (add all three student IDs before submitting)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: false
    latex_engine: xelatex
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.5)
```



## Loading Packages

```{r packages}
library(atrrr)       # Bluesky API access
library(tm)          # corpus and document-term matrices
library(SnowballC)   # stemming
library(wordcloud)   # word cloud visualisation
library(ggplot2)     # visualisations
library(dplyr)       # data wrangling
library(cluster)     # silhouette analysis
library(effsize)     # Cliff's delta effect size
library(knitr)       # tables
```

## Authenticating with Bluesky

```{r auth}
bsky_app_pw = "vc7d-2zda-wr7q-oyb3"
auth(user = "trungphanm.bsky.social", password = bsky_app_pw)
```

# Data

## Collecting Posts

The same monthly-window strategy as the exploratory script is used: each phrase is searched
separately for each month, because Bluesky search accepts one plain phrase per call. The result
is cached to `AI_bluesky.RData` after the first real download, so re-knitting this document does
not repeatedly hit the API. Delete that file and re-knit to trigger a fresh live collection.

```{r collect, eval=FALSE}
# WARNING: Bluesky content is uncontrolled and may contain offensive language.
queries = c("AI", "artificial intelligence", "generative AI", "ChatGPT")

# Twelve most recent full months at the time of writing.
months = seq(as.Date("2025-09-01"), as.Date("2026-09-01"), by = "month")
per_call = 400

all_posts = NULL
for (q in queries) {
  for (i in 1:(length(months) - 1)) {
    res = tryCatch(
      search_skeet(q, sort = "latest",
                   since = as.character(months[i]),
                   until = as.character(months[i + 1]),
                   limit = per_call),
      error = function(e) NULL
    )
    if (!is.null(res) && nrow(res) > 0) {
      res$query = q
      res$window = format(months[i], "%Y-%m")
      all_posts = rbind(all_posts, res)
    }
  }
}

# A post can match more than one phrase.
all_posts = all_posts[!duplicated(all_posts$uri), ]

# Keep English posts that contain text.
is_en = sapply(all_posts$langs, function(x) isTRUE(any(unlist(x) == "en")))
has_text = nchar(trimws(all_posts$text)) > 0
all_posts = all_posts[is_en & has_text, ]

save(all_posts, file = "AI_bluesky.RData")
```

```{r load-data}
data_file = "AI_bluesky.RData"
if (!file.exists(data_file)) {
  stop("AI_bluesky.RData not found. Run the collect chunk above (set eval=TRUE) first.")
}
load(data_file)

nrow(all_posts)
table(all_posts$query)
```

This gives `r nrow(all_posts)` English-language posts collected live from Bluesky across four
AI-related search phrases and twelve months, before any further cleaning.

# 2.2 Hypothesis Testing

## Significance Level

```{r alpha}
alpha = 0.05
```

## Hypothesis 1: ChatGPT and Generative AI

### Stating Keywords and Hypothesis

```{r h1-setup}
keyword_A = "chatgpt"
keyword_B = "generative"

# H0: mean likes of posts with keyword A == mean likes of posts with keyword B
# H1: the mean likes differ
```

### Filtering Data

```{r h1-filter}
data_keyword_A = all_posts %>% filter(grepl(keyword_A, text, ignore.case = TRUE))
data_keyword_B = all_posts %>% filter(grepl(keyword_B, text, ignore.case = TRUE))

nrow(data_keyword_A)
nrow(data_keyword_B)

cat("The number of likes with keyword A:", sum(data_keyword_A$like_count), "\n",
    "The number of likes with keyword B:", sum(data_keyword_B$like_count))
```

### Diagnostic Plots

```{r h1-diagnostics}
par(mfrow = c(1, 2))
qqnorm(data_keyword_A$like_count, main = paste("Q-Q Plot of Likes for", keyword_A))
qqline(data_keyword_A$like_count, col = "red")
qqnorm(data_keyword_B$like_count, main = paste("Q-Q Plot of Likes for", keyword_B))
qqline(data_keyword_B$like_count, col = "red")
par(mfrow = c(1, 1))

boxplot(data_keyword_A$like_count, data_keyword_B$like_count,
        names = c(keyword_A, keyword_B),
        main = "Boxplot of Likes", ylab = "Number of Likes")
```

### Two-Sample t-test

```{r h1-test}
t_test_h1 = t.test(data_keyword_A$like_count, data_keyword_B$like_count,
                   alternative = "two.sided", var.equal = FALSE)
t_test_h1

t_test_h1$p.value < alpha
```

## Hypothesis 2: Outrage and Enthusiasm

### Stating Keywords and Hypothesis

```{r h2-setup}
backlash_pattern = "\\b(slop|stolen|theft|scam)"
enthusiasm_pattern = "\\b(amazing|exciting|helpful|useful)"

# H0: mean likes of backlash posts == mean likes of enthusiasm posts
# H1: the mean likes differ
```

### Filtering Data

```{r h2-filter}
data_backlash = all_posts %>% filter(grepl(backlash_pattern, text, ignore.case = TRUE))
data_enthusiasm = all_posts %>% filter(grepl(enthusiasm_pattern, text, ignore.case = TRUE))

nrow(data_backlash)
nrow(data_enthusiasm)

cat("The number of likes on backlash posts:", sum(data_backlash$like_count), "\n",
    "The number of likes on enthusiasm posts:", sum(data_enthusiasm$like_count))
```

### Diagnostic Plots

```{r h2-diagnostics}
par(mfrow = c(1, 2))
qqnorm(data_backlash$like_count, main = "Q-Q Plot of Likes for Backlash")
qqline(data_backlash$like_count, col = "red")
qqnorm(data_enthusiasm$like_count, main = "Q-Q Plot of Likes for Enthusiasm")
qqline(data_enthusiasm$like_count, col = "red")
par(mfrow = c(1, 1))

boxplot(data_backlash$like_count, data_enthusiasm$like_count,
        names = c("Backlash", "Enthusiasm"),
        main = "Boxplot of Likes", ylab = "Number of Likes")
```

### Two-Sample t-test

```{r h2-test}
t_test_h2 = t.test(data_backlash$like_count, data_enthusiasm$like_count,
                   alternative = "two.sided", var.equal = FALSE)
t_test_h2

t_test_h2$p.value < alpha
```

## Hypothesis 3: Jobs and Art

### Stating Keywords and Hypothesis

```{r h3-setup}
# trailing boundary keeps "art" from matching "artificial"
job_pattern = "\\b(job|jobs|layoffs?|laid off|workers?|unemploy\\w*)\\b"
art_pattern = "\\b(art|arts|artists?|artwork|creatives?)\\b"

# H0: mean reposts of job posts == mean reposts of art posts
# H1: the mean reposts differ
```

### Filtering Data

```{r h3-filter}
data_job = all_posts %>% filter(grepl(job_pattern, text, ignore.case = TRUE))
data_art = all_posts %>% filter(grepl(art_pattern, text, ignore.case = TRUE))

nrow(data_job)
nrow(data_art)

cat("The number of reposts of job posts:", sum(data_job$repost_count), "\n",
    "The number of reposts of art posts:", sum(data_art$repost_count))
```

### Diagnostic Plots

```{r h3-diagnostics}
par(mfrow = c(1, 2))
qqnorm(data_job$repost_count, main = "Q-Q Plot of Reposts for Jobs")
qqline(data_job$repost_count, col = "red")
qqnorm(data_art$repost_count, main = "Q-Q Plot of Reposts for Art")
qqline(data_art$repost_count, col = "red")
par(mfrow = c(1, 1))

boxplot(data_job$repost_count, data_art$repost_count,
        names = c("Jobs", "Art"),
        main = "Boxplot of Reposts", ylab = "Number of Reposts")
```

### Two-Sample t-test

```{r h3-test}
t_test_h3 = t.test(data_job$repost_count, data_art$repost_count,
                   alternative = "two.sided", var.equal = FALSE)
t_test_h3

t_test_h3$p.value < alpha
```

## Hypothesis 4: Early and Late Period

### Stating Periods and Hypothesis

```{r h4-setup}
period_months = sort(unique(all_posts$window))
early_months = head(period_months, floor(length(period_months) / 2))

period_months
early_months

# H0: mean likes of early-period posts == mean likes of late-period posts
# H1: the mean likes differ
# recent posts have had less time to collect likes
```

### Filtering Data

```{r h4-filter}
data_early = all_posts %>% filter(window %in% early_months)
data_late = all_posts %>% filter(!(window %in% early_months))

nrow(data_early)
nrow(data_late)

cat("The number of likes in the early period:", sum(data_early$like_count), "\n",
    "The number of likes in the late period:", sum(data_late$like_count))
```

### Diagnostic Plots

```{r h4-diagnostics}
par(mfrow = c(1, 2))
qqnorm(data_early$like_count, main = "Q-Q Plot of Likes, Early Period")
qqline(data_early$like_count, col = "red")
qqnorm(data_late$like_count, main = "Q-Q Plot of Likes, Late Period")
qqline(data_late$like_count, col = "red")
par(mfrow = c(1, 1))

boxplot(data_early$like_count, data_late$like_count,
        names = c("Early", "Late"),
        main = "Boxplot of Likes", ylab = "Number of Likes")
```

### Two-Sample t-test

```{r h4-test}
t_test_h4 = t.test(data_early$like_count, data_late$like_count,
                   alternative = "two.sided", var.equal = FALSE)
t_test_h4

t_test_h4$p.value < alpha
```

## Hypothesis 5: Long and Short Posts

### Stating Length Groups and Hypothesis

```{r h5-setup}
posts_length = all_posts %>%
  mutate(word_count = lengths(strsplit(trimws(text), "\\s+")))

median_words = median(posts_length$word_count)
median_words

# H0: mean likes of posts above the median length == mean likes of posts at or below it
# H1: the mean likes differ
```

### Filtering Data

```{r h5-filter}
data_long = posts_length %>% filter(word_count > median_words)
data_short = posts_length %>% filter(word_count <= median_words)

nrow(data_long)
nrow(data_short)

cat("The number of likes on long posts:", sum(data_long$like_count), "\n",
    "The number of likes on short posts:", sum(data_short$like_count))
```

### Diagnostic Plots

```{r h5-diagnostics}
par(mfrow = c(1, 2))
qqnorm(data_long$like_count, main = "Q-Q Plot of Likes, Long Posts")
qqline(data_long$like_count, col = "red")
qqnorm(data_short$like_count, main = "Q-Q Plot of Likes, Short Posts")
qqline(data_short$like_count, col = "red")
par(mfrow = c(1, 1))

boxplot(data_long$like_count, data_short$like_count,
        names = c("Long", "Short"),
        main = "Boxplot of Likes", ylab = "Number of Likes")
```

### Two-Sample t-test

```{r h5-test}
t_test_h5 = t.test(data_long$like_count, data_short$like_count,
                   alternative = "two.sided", var.equal = FALSE)
t_test_h5

t_test_h5$p.value < alpha
```

## Summary of Tests

```{r hypothesis-summary}
tests = list(t_test_h1, t_test_h2, t_test_h3, t_test_h4, t_test_h5)

hypothesis_summary = data.frame(
  hypothesis = c("chatgpt vs generative", "backlash vs enthusiasm", "jobs vs art",
                 "early vs late", "long vs short"),
  measure = c("likes", "likes", "reposts", "likes", "likes"),
  n_A = c(nrow(data_keyword_A), nrow(data_backlash), nrow(data_job),
          nrow(data_early), nrow(data_long)),
  n_B = c(nrow(data_keyword_B), nrow(data_enthusiasm), nrow(data_art),
          nrow(data_late), nrow(data_short)),
  mean_A = sapply(tests, function(t) t$estimate[1]),
  mean_B = sapply(tests, function(t) t$estimate[2]),
  t_statistic = sapply(tests, function(t) t$statistic),
  p_value = sapply(tests, function(t) t$p.value))

hypothesis_summary$reject_H0 = hypothesis_summary$p_value < alpha
hypothesis_summary
```

# 2.3 Text/Content Analysis and Visualisation

## Removing Repeated Text

```{r dedupe}
posts_clean = all_posts[!duplicated(all_posts$text), ]

# posting month: created_at where present, otherwise the search window
posts_clean$month = if ("created_at" %in% names(posts_clean)) {
  substr(posts_clean$created_at, 1, 7)
} else {
  posts_clean$window
}

nrow(all_posts) - nrow(posts_clean)
nrow(posts_clean)
```

## Creating the Corpus

```{r corpus}
posts.text = posts_clean$text
index = posts_clean$query
post_month = posts_clean$month

posts.corpus = Corpus(VectorSource(posts.text))
posts.corpus = tm_map(posts.corpus, function(x) iconv(x, to = "ASCII", sub = " "))
```

## Pre-processing the Corpus

```{r preprocess}
remove_urls = content_transformer(function(x) {
  gsub("https?://\\S+|www\\.\\S+", " ", x, perl = TRUE)
})
posts.corpus = tm_map(posts.corpus, remove_urls)

custom_words = c("ai", "artificial", "intelligence", "amp",
                 "dont", "im", "ive", "youre", "theyre", "cant", "wont", "didnt", "isnt",
                 "just", "like", "really", "know", "get", "got", "one", "thing",
                 "say", "said", "also", "much", "many", "even", "still", "going", "gonna")

posts.corpus = tm_map(posts.corpus, removeNumbers)
posts.corpus = tm_map(posts.corpus, removePunctuation)
posts.corpus = tm_map(posts.corpus, stripWhitespace)
posts.corpus = tm_map(posts.corpus, tolower)
posts.corpus = tm_map(posts.corpus, removeWords, c(stopwords("english"), custom_words))
posts.corpus = tm_map(posts.corpus, stemDocument)
```

## Creating the Document-Term Matrix

```{r dtm}
posts.dtm = DocumentTermMatrix(posts.corpus)
dim(posts.dtm)

# terms in fewer than 1% of posts removed so the matrix fits in memory
posts.dtm = removeSparseTerms(posts.dtm, 0.99)
posts.matrix = as.matrix(posts.dtm)

# documents left empty by cleaning removed
empties = which(rowSums(posts.matrix) == 0)
if (length(empties) > 0) {
  posts.matrix = posts.matrix[-empties, ]
  posts.dtm = posts.dtm[-empties, ]
  index = index[-empties]
  post_month = post_month[-empties]
}

dim(posts.matrix)
```

## Finding the Top 20 Terms

```{r top-20}
w = colSums(posts.matrix)
o = order(w, decreasing = TRUE)[1:20]

top_20_words = colnames(posts.matrix)[o]
top_20_words
```

## TF-IDF Weighting

```{r builtin-tfidf}
posts.tdm = t(posts.dtm)
posts.wtdm = weightTfIdf(posts.tdm)
posts.weighted.tdm = as.matrix(posts.wtdm)
dim(posts.weighted.tdm)

sums = rowSums(posts.weighted.tdm)
o = order(sums, decreasing = TRUE)[1:20]

TFIDF_top20_words = names(sums)[o]
TFIDF_top20_words
```

## Comparing Frequency and TF-IDF Results

```{r compare-rankings}
data.frame(top_20_words, TFIDF_top20_words)
```

## Unweighted Word Cloud

```{r wordcloud-unweighted, fig.height=5.5}
freqs = rowSums(as.matrix(posts.tdm))
freqs = freqs[!is.na(freqs)]

wordcloud(names(freqs), freqs,
          random.order = FALSE,
          min.freq = 6,
          max.words = 200,
          colors = brewer.pal(8, "Dark2"),
          scale = c(2, .5))
title("Unweighted M (TDM)")
```

## TF-IDF Weighted Word Cloud

```{r wordcloud-weighted, fig.height=5.5}
freqsw = rowSums(posts.weighted.tdm)

wordcloud(names(freqsw), freqsw,
          random.order = FALSE,
          min.freq = 6,
          max.words = 200,
          colors = brewer.pal(8, "Dark2"),
          scale = c(1, .5))
title("TF-IDF Weighted T (TDM)")
```

## Word Clouds by Search Phrase

```{r wordcloud-by-phrase, fig.height=5.5}
data_list = split(posts_clean$text, posts_clean$query)

j = 1
for (i in data_list) {
  corpus = Corpus(VectorSource(i))
  corpus = tm_map(corpus, content_transformer(function(x) iconv(x, to = "ASCII", sub = " ")))
  corpus = tm_map(corpus, remove_urls)
  corpus = tm_map(corpus, removeNumbers)
  corpus = tm_map(corpus, removePunctuation)
  corpus = tm_map(corpus, stripWhitespace)
  corpus = tm_map(corpus, tolower)
  corpus = tm_map(corpus, removeWords, c(stopwords("en"), custom_words))
  corpus = tm_map(corpus, stemDocument)

  tdm = TermDocumentMatrix(corpus)
  # sparse terms removed so each phrase's matrix fits in memory
  tdm = removeSparseTerms(tdm, 0.99)
  tdm_matrix = as.matrix(tdm)

  freqs = rowSums(tdm_matrix)
  freqs = freqs[!is.na(freqs)]

  wordcloud(names(freqs), freqs,
            random.order = FALSE,
            min.freq = 3,
            max.words = 100,
            colors = brewer.pal(8, "Dark2"))
  mtext(names(data_list)[j], side = 3, line = 1, cex = 1.5)

  j = j + 1
}
```

# 2.4 Clustering

## Building the Weighted Document-Term Matrix

```{r cluster-matrix}
set.seed(3020)

# a sample keeps the N by N distance matrix tractable
max_cluster_posts = 3000
keep = sample(ncol(posts.weighted.tdm), min(max_cluster_posts, ncol(posts.weighted.tdm)))

posts.cmatrix = t(posts.weighted.tdm[, keep])
index.clust = index[keep]
month.clust = post_month[keep]

empties = which(rowSums(abs(posts.cmatrix)) == 0)
if (length(empties) > 0) {
  posts.cmatrix = posts.cmatrix[-empties, ]
  index.clust = index.clust[-empties]
  month.clust = month.clust[-empties]
}

dim(posts.cmatrix)
```

## K-Means Clustering

```{r kmeans-one}
number.of.clusters = 1
K = kmeans(posts.cmatrix, number.of.clusters)

K$tot.withinss
K$totss
```

## Applying the Elbow Method

```{r elbow-euclidean}
n = 15
SSW = rep(0, n)
for (a in 1:n) {
  K = kmeans(posts.cmatrix, a, nstart = 10)
  SSW[a] = K$tot.withinss
}

plot(1:15, SSW, type = "b", xaxt = "n", xlab = "k")
axis(1, at = 1:15)
```

## Calculating Cosine Distance

```{r cosine-distance}
norm.posts.matrix = diag(
  1/sqrt(rowSums(posts.cmatrix^2))
) %*% posts.cmatrix

D = dist(norm.posts.matrix, method = "euclidean")^2/2
```

## Projecting into Euclidean Space

```{r mds}
N = dim(posts.cmatrix)[1]
N

mds.posts.matrix = cmdscale(D, k = min(N - 1, ncol(posts.cmatrix)))

dim(as.matrix(D))
dim(mds.posts.matrix)
```

## Selecting the Number of Clusters

```{r elbow-silhouette}
n = 15
SSW = rep(0, n)
sil_score = rep(NA, n)

for (a in 1:n) {
  K = kmeans(mds.posts.matrix, a, nstart = 20)
  SSW[a] = K$tot.withinss
  if (a > 1) sil_score[a] = mean(silhouette(K$cluster, D)[, 3])
}

par(mfrow = c(1, 2))
plot(1:15, SSW, type = "b", xaxt = "n", xlab = "k")
axis(1, at = 1:15)
plot(1:15, sil_score, type = "b", xaxt = "n", xlab = "k")
axis(1, at = 1:15)
par(mfrow = c(1, 1))

data.frame(k = 1:n, WSS = SSW, Silhouette_Score = sil_score)
```

## K-Means with Cosine Distance

```{r kmeans-cosine}
# set from the elbow and silhouette plots above
number.of.clusters = 5

K = kmeans(mds.posts.matrix, number.of.clusters, nstart = 20)
summary(K)
table(K$cluster)

cluster_sizes = table(K$cluster)
min(cluster_sizes) / max(cluster_sizes)
```

## Visualising the Clusters

```{r plot-clusters}
mds2.posts.matrix = cmdscale(D, k = 2)
plot(mds2.posts.matrix, col = K$cluster)
```

## Examining Cluster Contents

```{r cluster-contents}
for (cluster.number in 1:number.of.clusters) {
  clusterpostsId = which(K$cluster == cluster.number)
  clusterposts = posts.cmatrix[clusterpostsId, , drop = FALSE]
  clusterTermWeight = colMeans(clusterposts)
  o = order(clusterTermWeight, decreasing = TRUE)[1:10]
  cat("Cluster", cluster.number, "(", length(clusterpostsId), "posts):",
      colnames(posts.cmatrix)[o], "\n")
}
```

## Comparing Clusters with Search Phrases

```{r cluster-vs-phrase}
table(K$cluster, index.clust)
```

## Hierarchical Clustering

```{r hierarchical, fig.height=5}
H = hclust(D, method = "complete")
plot(H, labels = FALSE)

groups = cutree(H, k = number.of.clusters)
table(groups)
table(groups, index.clust)
```