---
title: "The AI Conversation on Bluesky"
subtitle: "COMP3020 Social Web Analytics -- Hypothesis Testing, Text Analysis and Clustering"
author: "Trung, Thu and Christan (add all three student IDs before submitting)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: false
    latex_engine: xelatex
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.5)
```



## Loading Packages

```{r packages}
library(atrrr)       # Bluesky API access
library(tm)          # corpus and document-term matrices
library(SnowballC)   # stemming
library(wordcloud)   # word cloud visualisation
library(ggplot2)     # visualisations
library(dplyr)       # data wrangling
library(cluster)     # silhouette analysis
library(effsize)     # Cliff's delta effect size
library(knitr)       # tables
```

## Authenticating with Bluesky

```{r auth}
bsky_app_pw = "vc7d-2zda-wr7q-oyb3"
auth(user = "trungphanm.bsky.social", password = bsky_app_pw)
```

# Data

## Collecting Posts

The same monthly-window strategy as the exploratory script is used: each phrase is searched
separately for each month, because Bluesky search accepts one plain phrase per call. The result
is cached to `AI_bluesky.RData` after the first real download, so re-knitting this document does
not repeatedly hit the API. Delete that file and re-knit to trigger a fresh live collection.

```{r collect, eval=FALSE}
# WARNING: Bluesky content is uncontrolled and may contain offensive language.
queries = c("AI", "artificial intelligence", "generative AI", "ChatGPT")

# Twelve most recent full months at the time of writing.
months = seq(as.Date("2025-09-01"), as.Date("2026-09-01"), by = "month")
per_call = 400

all_posts = NULL
for (q in queries) {
  for (i in 1:(length(months) - 1)) {
    res = tryCatch(
      search_skeet(q, sort = "latest",
                   since = as.character(months[i]),
                   until = as.character(months[i + 1]),
                   limit = per_call),
      error = function(e) NULL
    )
    if (!is.null(res) && nrow(res) > 0) {
      res$query = q
      res$window = format(months[i], "%Y-%m")
      all_posts = rbind(all_posts, res)
    }
  }
}

# A post can match more than one phrase.
all_posts = all_posts[!duplicated(all_posts$uri), ]

# Keep English posts that contain text.
is_en = sapply(all_posts$langs, function(x) isTRUE(any(unlist(x) == "en")))
has_text = nchar(trimws(all_posts$text)) > 0
all_posts = all_posts[is_en & has_text, ]

save(all_posts, file = "AI_bluesky.RData")
```

```{r load-data}
data_file = "AI_bluesky.RData"
if (!file.exists(data_file)) {
  stop("AI_bluesky.RData not found. Run the collect chunk above (set eval=TRUE) first.")
}
load(data_file)

nrow(all_posts)
table(all_posts$query)
```

This gives `r nrow(all_posts)` English-language posts collected live from Bluesky across four
AI-related search phrases and twelve months, before any further cleaning.

# 2.2 Hypothesis Testing

## Research Question

Two of the search phrases point at very different kinds of talk: **"ChatGPT"** names a specific,
branded product, while **"AI"**, **"artificial intelligence"** and **"generative AI"** are used
for more general discussion of the technology. A natural question for the overall investigation
into what the AI conversation on Bluesky looks like is whether these two kinds of post are
received differently by the audience.

> **Research question:** Do posts that name a specific AI product (ChatGPT) receive a different
> level of engagement, measured by the number of likes, than posts that discuss AI in more
> general terms?

## Hypotheses

- $H_0$: The distribution of `like_count` is the same for ChatGPT-query posts and general-AI-query
  posts (no location shift between the two groups).
- $H_1$: The distribution of `like_count` differs between ChatGPT-query posts and general-AI-query
  posts (two-sided).

## Choosing the Test

```{r hyp-groups}
all_posts$group = ifelse(all_posts$query == "ChatGPT", "ChatGPT", "General AI terms")
table(all_posts$group)
```

```{r hyp-shape}
ggplot(all_posts, aes(x = like_count)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(trans = "log1p", breaks = c(0, 1, 5, 20, 100, 500, 2000)) +
  facet_wrap(~ group, ncol = 1, scales = "free_y") +
  labs(title = "Distribution of likes is heavily right-skewed with a large mass at zero",
       x = "Likes (log1p scale)", y = "Number of posts")
```

`like_count` is a non-negative count with a median of `r median(all_posts$like_count)`, a mean of
`r round(mean(all_posts$like_count), 1)`, and a maximum of `r max(all_posts$like_count)`. Most
posts get very few or no likes, while a small number go viral, so the distribution is far from
normal in both groups (the histogram above is on a log scale and still shows a long tail). An
independent-samples $t$-test assumes approximately normal, equal-variance samples, which does not
hold here. The **Wilcoxon rank-sum test (Mann-Whitney U test)**, covered in COMP3020 as the
non-parametric alternative for comparing two independent groups, only requires the two
distributions to have the same shape up to a location shift and is robust to skew and outliers,
so it is the appropriate test for this comparison.

## Running the Test

```{r hyp-summary}
summary_tab = all_posts %>%
  group_by(group) %>%
  summarise(n = n(),
            median_likes = median(like_count),
            mean_likes = round(mean(like_count), 2),
            IQR_likes = IQR(like_count))
kable(summary_tab, caption = "Engagement summary by query group")
```

```{r hyp-boxplot}
ggplot(all_posts, aes(x = group, y = like_count + 1)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Likes by query group (log scale, +1 to show zeros)",
       x = NULL, y = "Likes + 1")
```

```{r hyp-test}
wt = wilcox.test(like_count ~ group, data = all_posts, exact = FALSE, conf.int = TRUE)
wt

cd = cliff.delta(like_count ~ group, data = all_posts)
cd
```

## Interpretation

The test is significant (W = `r format(wt$statistic, big.mark = ",")`,
p = `r format.pval(wt$p.value, digits = 3)`), so $H_0$ is rejected: the two groups do not have the
same distribution of likes. ChatGPT-query posts have a slightly higher median and mean like count
than general-AI-query posts (Table above).

However, the effect size tells a more useful story than the p-value alone. Cliff's delta is
`r round(cd$estimate, 3)`, which the `effsize` package itself labels **`r cd$magnitude`**. With
`r nrow(all_posts)` posts, even a tiny, practically unimportant shift in the distribution becomes
statistically detectable. So the honest reading of this result is: there **is** a real difference
in typical engagement between posts naming a specific AI product and posts discussing AI in
general terms, but the size of that difference is small enough that query type alone is a weak
predictor of how much engagement an individual post gets. This matters for the overall
investigation because it means engagement should not be used as a proxy for which *topic* is
"winning" the conversation other factors (author reach, timing, whether a post is a reply,
etc.) are doing most of the work, and the thematic content explored in the next two sections is a
better lens on the conversation than raw like counts.

# 2.3 Text/Content Analysis and Visualisation

## Cleaning the Data

The collection step already restricted the data to English-language posts with non-empty text and
removed exact duplicate URIs (a post can match more than one search phrase). Two further issues
need addressing before the text can be analysed:

```{r clean-dupes}
n_before = nrow(all_posts)
all_posts = all_posts[!duplicated(all_posts$text), ]
n_after_dedupe = nrow(all_posts)
```

- **Repeated text.** `r n_before - n_after_dedupe` posts share exactly the same text as an
  earlier post (retweet-style reposts, quoted boilerplate, or bot accounts posting identical
  content). These are removed so that frequent-word counts and clusters reflect distinct
  contributions rather than the same message counted many times.
- **Missing engagement values.** The engagement counters (`like_count`, `repost_count`,
  `reply_count`, `quote_count`) come back as `0`, never `NA`, from the API, so no imputation is
  needed there; this is confirmed rather than assumed:

```{r clean-na-check}
colSums(is.na(all_posts[, c("like_count", "repost_count", "reply_count", "quote_count")]))
```

That leaves `r n_after_dedupe` posts for the text analysis.

## Building the Corpus

```{r corpus}
posts.text = all_posts$text
posts.corpus = Corpus(VectorSource(posts.text))

# Convert to ASCII to reduce encoding inconsistencies (emoji, curly quotes, etc.).
posts.corpus = tm_map(posts.corpus, function(x) iconv(x, to = "ASCII", sub = " "))
```

## Pre-processing

```{r preprocess}
# Remove links first so domain fragments do not survive as terms.
remove_urls = content_transformer(function(x) {
  gsub("https?://\\S+|www\\.\\S+", " ", x, perl = TRUE)
})
posts.corpus = tm_map(posts.corpus, remove_urls)

posts.corpus = tm_map(posts.corpus, removeNumbers)
posts.corpus = tm_map(posts.corpus, removePunctuation)
posts.corpus = tm_map(posts.corpus, stripWhitespace)
posts.corpus = tm_map(posts.corpus, tolower)

# Search terms are removed because they are common by construction, not by content.
# Punctuation is stripped before stopwords are removed, so contractions such as
# "don't" become "dont" and survive the standard stopword list; those remnants, plus
# a handful of generic filler words identified after inspecting an initial run of the
# top-term list below, are added so the remaining terms carry more topical content.
custom_words = c("ai", "artificial", "intelligence", "amp",
                   "dont", "im", "ive", "youre", "theyre", "cant", "wont", "didnt", "isnt",
                   "just", "like", "really", "know", "get", "got", "one", "thing",
                   "say", "said", "also", "much", "many", "even", "still", "going", "gonna")
posts.corpus = tm_map(posts.corpus, removeWords, c(stopwords("english"), custom_words))

posts.corpus = tm_map(posts.corpus, stemDocument)
```

## Representing the Text

```{r dtm}
posts.dtm = DocumentTermMatrix(posts.corpus)
dim(posts.dtm)

# Some short posts (e.g. a bare "AI" mention, or one that is mostly emoji/mentions)
# have no terms left after cleaning; these carry no distance/frequency information
# and are unsuitable records for the analyses below, so they are dropped here.
doc_lengths = slam::row_sums(posts.dtm)
empty_docs = which(doc_lengths == 0)
length(empty_docs)

posts.dtm = posts.dtm[doc_lengths > 0, ]
all_posts = all_posts[doc_lengths > 0, ]

# With this many posts the full matrix cannot be densified, so the sparsest terms
# are dropped first. 0.99 keeps terms appearing in at least one percent of posts.
posts.dtm = removeSparseTerms(posts.dtm, 0.99)
posts.matrix = as.matrix(posts.dtm)
dim(posts.matrix)
```

Bag-of-words counts are an appropriate representation here because the goal in this section is
summarising *which words* dominate the conversation, not the grammar connecting them; word order
is deliberately discarded. `r length(empty_docs)` posts were dropped as unsuitable records because
cleaning left them with no analysable content, and the matrix is trimmed to terms that appear in
at least 1% of posts, since extremely rare terms cannot support a summary at this scale and would
make the matrix impossible to work with densely.

## Textual Characteristics

```{r post-length}
word_counts = sapply(strsplit(trimws(posts.text), "\\s+"), length)
summary(word_counts)
```

```{r viz-length, fig.cap="Distribution of post length in raw words, before cleaning"}
ggplot(data.frame(words = word_counts), aes(x = words)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  labs(title = "Most AI-related Bluesky posts are short",
       x = "Words per post", y = "Number of posts")
```

Posts are short: the median is `r median(word_counts)` words and the third quartile is only
`r quantile(word_counts, 0.75)` words, consistent with Bluesky's character limit and with a
conversation carried in quick reactions rather than long-form writing. This matters for the
clustering in the next section, since short documents give k-means less signal to separate on.

## Frequent Words

```{r top-terms}
w = colSums(posts.matrix)
o = order(w, decreasing = TRUE)[1:20]
top_terms = data.frame(term = colnames(posts.matrix)[o], frequency = w[o])
kable(top_terms, row.names = FALSE, caption = "Top 20 terms by raw frequency")
```

```{r viz-topterms, fig.cap="Most frequent terms across the cleaned corpus"}
ggplot(top_terms, aes(x = reorder(term, frequency), y = frequency)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 20 most frequent terms", x = NULL, y = "Number of posts containing term")
```

```{r wordcloud-raw, fig.height=5.5, fig.cap="Word cloud of unweighted term frequencies"}
freqs = colSums(posts.matrix)
wordcloud(names(freqs), freqs,
          random.order = FALSE, min.freq = 20, max.words = 150,
          colors = brewer.pal(8, "Dark2"), scale = c(3, 0.5))
```

## Interpreting the Patterns

Alongside product/tool terms (`chatgpt`, `openai`, `gemini` not shown in the top 20 but present
further down the table), the top terms split into two broad groups: action/creation words
(`use`, `generat`, `creat`, `make`, `work`, `write`) and people/impact words (`peopl`, `human`,
`job`, `compani`). That split already hints at the two themes recovered more formally in the
clustering below -- AI as a *tool being used* versus AI as a *force acting on people and
institutions* -- and it is the first concrete evidence, ahead of the clustering, that the
conversation is not a single homogeneous topic. `generat` sits near the top largely because it is
the stem of "generative", reflecting the search phrase, and should be read with that in mind
rather than as a sign that people talk about "generating" things unprompted.

# 2.4 Clustering

## Representation and Distance

Clustering reuses the document-term matrix built above, but re-weights it with **TF-IDF** rather
than raw counts: raw frequency rewards common, low-information stems (which is exactly what the
frequency table above is for), whereas TF-IDF down-weights terms that occur in most posts and
up-weights terms that are distinctive of a smaller set of posts, which is what actually separates
topics.

```{r tfidf}
posts.tdm = t(posts.dtm)
posts.wtdm = weightTfIdf(posts.tdm)
```

Pairwise-distance clustering needs an $N \times N$ distance matrix, so cost grows with the square
of the number of posts; a random sample keeps the computation tractable while still giving a
representative cross-section of the twelve-month collection.

```{r cluster-sample}
set.seed(3020)
max_cluster_posts = 1500
keep = sample(ncol(posts.wtdm), min(max_cluster_posts, ncol(posts.wtdm)))

posts.wmatrix = t(as.matrix(posts.wtdm[, keep]))   # documents in rows
sample_posts = all_posts[keep, ]

empties = which(rowSums(abs(posts.wmatrix)) == 0)
if (length(empties) > 0) {
  posts.wmatrix = posts.wmatrix[-empties, ]
  sample_posts = sample_posts[-empties, ]
}
dim(posts.wmatrix)
```

Documents are TF-IDF vectors of *proportional* term usage, so two posts about the same topic
should count as close regardless of how many words either one happens to use; that is exactly
what **cosine distance** measures (the angle between vectors), whereas plain Euclidean distance on
raw vectors would also pick up differences in post length as if they were topical differences.
Cosine distance is obtained here by normalising each document vector to unit length and then
using Euclidean distance, which is mathematically equivalent to cosine distance on normalised
vectors; $k$-means itself only operates in Euclidean space, so classical MDS is used to project the
distance matrix into coordinates it can cluster.

```{r cosine-mds}
norm.posts.matrix = posts.wmatrix / sqrt(rowSums(posts.wmatrix^2))
D = dist(norm.posts.matrix, method = "euclidean")^2 / 2

N = nrow(posts.wmatrix)
mds.posts.matrix = cmdscale(D, k = min(100, N - 1))
dim(mds.posts.matrix)
```

## Choosing the Number of Clusters

A single diagnostic can be misleading, so two are used together: the within-cluster sum of squares
(elbow method) and the average silhouette width, which additionally checks that points are closer
to their own cluster than to the nearest neighbouring cluster rather than just checking that
overall spread has been reduced.

```{r elbow-silhouette}
n = 12
SSW = rep(0, n)
sil = rep(NA, n)
for (a in 1:n) {
  K_a = kmeans(mds.posts.matrix, a, nstart = 10)
  SSW[a] = K_a$tot.withinss
  if (a > 1) sil[a] = mean(silhouette(K_a$cluster, D)[, 3])
}

par(mfrow = c(1, 2))
plot(1:n, SSW, type = "b", xlab = "Number of clusters", ylab = "Within-cluster SS",
     main = "Elbow")
plot(2:n, sil[2:n], type = "b", xlab = "Number of clusters", ylab = "Mean silhouette width",
     main = "Silhouette")
```

The elbow flattens out from around six clusters onward, and the (uniformly low, as is typical for
short, noisy social-media text) silhouette width also firms up from six clusters onward rather
than continuing to climb only because more clusters mechanically shrink each one. Six was
confirmed, not just read off the plot, by inspecting the top terms and a sample of original posts
in each of the candidate clusterings around that value (below): fewer clusters merged clearly
different themes (e.g. product news and personal backlash into one group); more clusters mostly
split existing themes into near-duplicates rather than revealing new ones.

```{r kmeans}
number.of.clusters = 6
K = kmeans(mds.posts.matrix, number.of.clusters, nstart = 25)
table(K$cluster)
```

```{r viz-clusters, fig.cap="Sampled posts clustered by cosine distance, projected to two MDS dimensions"}
mds2.posts.matrix = cmdscale(D, k = 2)
cluster_df = data.frame(mds2.posts.matrix, cluster = factor(K$cluster))
names(cluster_df)[1:2] = c("dim1", "dim2")

ggplot(cluster_df, aes(x = dim1, y = dim2, colour = cluster)) +
  geom_point(alpha = 0.6, size = 1.2) +
  labs(title = "Sampled posts clustered by cosine distance",
       x = "MDS dimension 1", y = "MDS dimension 2", colour = "Cluster")
```

The MDS plot shows one large, diffuse cluster with several smaller, tighter clusters pulled out
around its edge -- consistent with a broad general conversation about AI with a few specific,
more tightly-worded sub-topics embedded in it.

## Cluster Themes

```{r cluster-terms}
top_per_cluster = 12
theme_terms = list()

for (cluster.number in 1:number.of.clusters) {
  ids = which(K$cluster == cluster.number)
  cluster_term_weight = colMeans(posts.wmatrix[ids, , drop = FALSE])
  o = order(cluster_term_weight, decreasing = TRUE)[1:top_per_cluster]
  stems = colnames(posts.wmatrix)[o]
  theme_terms[[paste0("cluster", cluster.number)]] = stems
  cat("Cluster", cluster.number, "(", length(ids), "posts):", stems, "\n")
}
```

```{r cluster-query-mix}
# Cross-checking clusters against the original search phrase gives an independent
# read on whether the clusters found in the text line up with something meaningful,
# rather than accepting the top-term lists at face value.
table(K$cluster, sample_posts$query)
```

```{r cluster-examples}
set.seed(3020)
for (cluster.number in 1:number.of.clusters) {
  ids = which(K$cluster == cluster.number)
  cat("--- Cluster", cluster.number, "example posts ---\n")
  print(sample_posts$text[sample(ids, min(2, length(ids)))])
  cat("\n")
}
```

Reading the top terms, the query mix, and the example posts together (rather than the term list
alone) gives six interpretable themes:

1. **Product/company news** -- dominated by `chatgpt`, `openai`, `gemini`; almost entirely posts
   collected under the "ChatGPT" phrase. This is the same group singled out in the hypothesis test
   above, and its tight, news-style language (headlines, company names) is a plausible reason that
   group's engagement differs slightly from the rest.
2. **Personal/emotional backlash** -- profanity and first-person language about being replaced or
   resentful of AI use; a smaller, emotionally distinct pocket of the conversation rather than
   neutral commentary.
3. **General AI commentary and news** -- the largest, most diffuse cluster (companies, technology,
   history, current events); this is the broad background conversation that the other, tighter
   clusters sit inside.
4. **Automated/bot content** -- short, templated posts starting with "Feed:" from RSS-style
   aggregation bots. This is not a topic at all but an artefact of the collection: some accounts
   auto-post every item from a news feed. It is a useful finding in its own right -- it shows that
   a small slice of "AI conversation" volume is not conversation at all, and a stricter cleaning
   pass for further work should consider filtering by author posting pattern, not just language and
   duplicate text.
5. **Creative industries and generative AI** -- artists, "content", games; concentrated in the
   "generative AI" phrase, discussing generative tools' effect on creative work specifically.
6. **Reflective/opinion posts** -- first-person musing about the future, writing, and long-running
   personal projects; a smaller cluster of think-piece-style posts rather than news or reaction.

The clustering therefore goes beyond simply separating "AI" posts into arbitrary bins: it recovers
a recognisable structure of product news, emotional reaction, general commentary, creative-industry
discussion, personal reflection, and -- usefully -- automated noise, which together describe the
shape of the AI conversation on Bluesky better than any single summary statistic could.
