---
title: "Finding the AI conversation on Bluesky"
subtitle: "COMP3020 Social Web Analytics: exploratory phase"
author: "Trung, Thu and Christan (add all three student IDs before submitting)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.5)
```

# Design

The topic is not fixed in advance. The collection is broad, the patterns come out of the
text, and the focus is chosen from what the patterns show.

- Phase 1: collect posts on AI and related terms, month by month, so the sample covers a
  period rather than a single moment.
- Phase 2: find patterns with word frequency, TF-IDF and clustering.
- Phase 3: name the themes the clusters reveal and label every post with one.
- Phase 4: track the themes over time, then choose the focus topic for the poster.

Only the methods needed for these phases are included. Network analysis and link analysis
follow once the topic is chosen.

# Setup

## Loading Packages

```{r packages}
# Functions for retrieving Bluesky posts.
library(atrrr)
```

## Authenticating with Bluesky

```{r auth}
# Authorize the API to load dataset
 bsky_app_pw <-"vc7d-2zda-wr7q-oyb3"
 auth(user = "trungphanm.bsky.social", password = bsky_app_pw)
```

## Checking the Search Arguments

```{r check-args}
# The monthly collection needs both since and until. Confirm your version has them.
args(search_skeet)
```

If `until` is not listed, remove it from the search call below and filter on the post
timestamp afterwards. The monthly windows will then be uneven, because `sort = "latest"`
returns the most recent posts overall rather than the most recent inside a window.

# Phase 1: Collecting Bluesky Posts

## Searching Month by Month

Each phrase is searched once per month. Bluesky search takes one plain phrase at a time, so
several phrases are used to reach the whole conversation rather than one corner of it.

The collection is switched off once the data is saved. Set `eval=TRUE` on this chunk and the
next one to collect again.

```{r collect, eval=FALSE}
# WARNING: Bluesky content is uncontrolled and may contain offensive language.
queries = c("AI", "artificial intelligence", "generative AI", "ChatGPT")

# Bluesky opened to the public in February 2024, so earlier windows return nothing.
months = seq(as.Date("2024-01-01"), as.Date("2026-10-01"), by = "month")
per_call = 500

all_posts = NULL

for (q in queries) {
  for (i in 1:(length(months) - 1)) {
    res = search_skeet(q,
                       sort = "latest",
                       since = as.character(months[i]),
                       until = as.character(months[i + 1]),
                       limit = per_call
                       )
    if (nrow(res) > 0) {
      res$query = q
      res$window = format(months[i], "%Y-%m")
      all_posts = rbind(all_posts, res)
    }
  }
}
```

## Filtering and Saving

```{r filter-save, eval=FALSE}
# A post can match more than one phrase.
all_posts = all_posts[!duplicated(all_posts$uri), ]

# Keep English posts that contain text.
is_en = sapply(all_posts$langs, function(x) isTRUE(any(unlist(x) == "en")))
has_text = nchar(trimws(all_posts$text)) > 0
all_posts = all_posts[is_en & has_text, ]

save(all_posts, file = "AI_bluesky.RData")
```

## Loading the Saved Data

```{r load-data}
load("AI_bluesky.RData")

nrow(all_posts)
names(all_posts)
```

## Checking the Collection

```{r check-collection}
# Posts per phrase and per window. Any cell at the limit was cut off.
table(all_posts$query)
table(all_posts$window)
```

# Phase 2: Preparing the Text

## Creating the Corpus

```{r corpus}
library(tm)
library(SnowballC)

posts.text = all_posts$text
posts.corpus = Corpus(VectorSource(posts.text))

# Convert to ASCII to reduce encoding inconsistencies.
posts.corpus = tm_map(posts.corpus, function(x) iconv(x, to = "ASCII", sub = " "))
```

## Pre-processing the Corpus

```{r preprocess}
# Remove links first so domain fragments do not survive as terms.
remove_urls = content_transformer(function(x) {
  gsub("https?://\\S+|www\\.\\S+", " ", x, perl = TRUE)
})
posts.corpus = tm_map(posts.corpus, remove_urls)

posts.corpus = tm_map(posts.corpus, removeNumbers)
posts.corpus = tm_map(posts.corpus, removePunctuation)
posts.corpus = tm_map(posts.corpus, stripWhitespace)
posts.corpus = tm_map(posts.corpus, tolower)

# Only the words present in every post by construction are removed. Product names
# such as chatgpt are kept, because which tools people name is part of the pattern.
custom_words = c("ai", "artificial", "intelligence", "amp")
posts.corpus = tm_map(posts.corpus, removeWords,
                      c(stopwords("english"), custom_words))

posts.corpus = tm_map(posts.corpus, stemDocument)
```

# Phase 2: Finding Patterns

## Creating the Document-Term Matrix

```{r dtm}
posts.dtm = DocumentTermMatrix(posts.corpus)
dim(posts.dtm)

# With this many posts the full matrix cannot be densified, so the sparsest terms
# are dropped first. 0.99 keeps terms appearing in at least one percent of posts.
posts.dtm = removeSparseTerms(posts.dtm, 0.99)
posts.matrix = as.matrix(posts.dtm)
dim(posts.matrix)
```

## Top Terms by Frequency

```{r top-terms}
w = colSums(posts.matrix)
o = order(w, decreasing = TRUE)[1:30]
colnames(posts.matrix)[o]
```

## Top Terms by TF-IDF

```{r tfidf}
# Transpose the reduced matrix rather than rebuilding it from the corpus.
posts.tdm = t(posts.dtm)

posts.wtdm = weightTfIdf(posts.tdm)
posts.weighted.tdm = as.matrix(posts.wtdm)

sums = rowSums(posts.weighted.tdm)
o = order(sums, decreasing = TRUE)[1:30]
names(sums)[o]
```

## Word Clouds

```{r wordcloud-raw, fig.height=5.5}
library(wordcloud)

freqs = rowSums(as.matrix(posts.tdm))

wordcloud(names(freqs), freqs,
          random.order = FALSE, min.freq = 20, max.words = 150,
          colors = brewer.pal(8, "Dark2"), scale = c(3, 0.5))
title("Unweighted term frequencies")
```

```{r wordcloud-weighted, fig.height=5.5}
freqsw = rowSums(posts.weighted.tdm)

wordcloud(names(freqsw), freqsw,
          random.order = FALSE, min.freq = 0, max.words = 150,
          colors = brewer.pal(8, "Dark2"), scale = c(2.5, 0.4))
title("TF-IDF weighted terms")
```

## Clustering a Sample of Posts

Cosine distance needs an N by N matrix and MDS decomposes it, so the cost grows with the
square of the post count. The frequency work above uses every post. The clustering uses a
random sample.

```{r cluster-sample}
set.seed(3020)

max_cluster_posts = 2000
keep = sample(ncol(posts.wtdm), min(max_cluster_posts, ncol(posts.wtdm)))

# Clustering functions expect documents in rows.
posts.wmatrix = t(as.matrix(posts.wtdm[, keep]))

# Cleaning leaves some posts with no terms; zero rows carry no distance information.
empties = which(rowSums(abs(posts.wmatrix)) == 0)
if (length(empties) > 0) posts.wmatrix = posts.wmatrix[-empties, ]

dim(posts.wmatrix)
```

```{r cosine-mds}
# Normalise each document vector to unit length, then use cosine distance.
norm.posts.matrix = posts.wmatrix / sqrt(rowSums(posts.wmatrix^2))
D = dist(norm.posts.matrix, method = "euclidean")^2 / 2

# k-means only supports Euclidean geometry, so MDS projects the cosine distances.
N = nrow(posts.wmatrix)
mds.posts.matrix = cmdscale(D, k = min(100, N - 1))
dim(mds.posts.matrix)
```

```{r elbow}
library(cluster)

n = 12
SSW = rep(0, n)
SIL = rep(NA, n)

for (a in 1:n) {
  K = kmeans(mds.posts.matrix, a, nstart = 10)
  SSW[a] = K$tot.withinss
  # Silhouette needs at least two clusters.
  if (a > 1) SIL[a] = mean(silhouette(K$cluster, D)[, 3])
}

par(mfrow = c(1, 2))
plot(1:n, SSW, type = "b",
     xlab = "Number of clusters", ylab = "Within-cluster sum of squares")
plot(1:n, SIL, type = "b",
     xlab = "Number of clusters", ylab = "Average silhouette width")
par(mfrow = c(1, 1))

data.frame(k = 1:n, SSW = round(SSW, 2), silhouette = round(SIL, 3))
```

```{r kmeans}
# Read the elbow from the plot above before fixing this value.
number.of.clusters = 5

K = kmeans(mds.posts.matrix, number.of.clusters, nstart = 20)
table(K$cluster)

mds2.posts.matrix = cmdscale(D, k = 2)
plot(mds2.posts.matrix, col = K$cluster, pch = 19, cex = 0.5,
     xlab = "MDS dimension 1", ylab = "MDS dimension 2",
     main = "Sampled posts clustered by cosine distance")
```

```{r cluster-terms}
# Average each term's weight across a cluster. The highest means describe it, and
# they become the starting dictionary for the themes.
top_per_cluster = 10
theme_terms = list()

for (cluster.number in 1:number.of.clusters) {
  clusterpostsId = which(K$cluster == cluster.number)
  clusterTermWeight = colMeans(posts.wmatrix[clusterpostsId, , drop = FALSE])
  o = order(clusterTermWeight, decreasing = TRUE)[1:top_per_cluster]
  stems = colnames(posts.wmatrix)[o]

  theme_terms[[paste0("cluster", cluster.number)]] = stems
  cat("Cluster", cluster.number, "(", length(clusterpostsId), "posts):", stems, "\n")
}
```

# Phase 3: Naming the Themes

The dictionary comes from the clusters above, so the themes are taken from the data rather
than assumed. Read the cluster terms, then give each cluster a name.

```{r theme-names}
# Rename once the cluster terms have been read, for example:
# names(theme_terms) = c("work", "art", "study", "tools", "policy")

# Stems shared by several clusters cannot separate themes, so drop them.
all_stems = unlist(theme_terms)
shared = names(which(table(all_stems) > 1))
theme_terms = lapply(theme_terms, function(x) setdiff(x, shared))

theme_terms
```

```{r themes}
# A post takes the first theme it matches, so the order of the list matters.
cleaned.text = sapply(posts.corpus, as.character)

theme = rep("other", length(cleaned.text))
for (nm in names(theme_terms)) {
  stems = theme_terms[[nm]]
  if (length(stems) == 0) next
  pattern = paste0("\\b(", paste(stems, collapse = "|"), ")")
  hit = grepl(pattern, cleaned.text)
  theme[hit & theme == "other"] = nm
}

table(theme)
```

# Phase 4: Tracking the Themes Over Time

## Posts per Month

```{r trend-volume}
# Column names differ between atrrr versions, so resolve the timestamp first.
stamp = NULL
if ("created_at" %in% names(all_posts)) stamp = all_posts$created_at
if (is.null(stamp) && "indexed_at" %in% names(all_posts)) stamp = all_posts$indexed_at
if (is.null(stamp) && "post_data" %in% names(all_posts)) {
  stamp = sapply(all_posts$post_data, function(x) {
    v = unlist(x)
    hit = grep("^[0-9]{4}-[0-9]{2}-[0-9]{2}T", v, value = TRUE)
    if (length(hit) > 0) hit[1] else NA_character_
  })
}

all_posts$month = if (is.null(stamp)) NA_character_ else
  format(as.Date(substr(stamp, 1, 10)), "%Y-%m")

# Fall back to the search window if no timestamp survived.
if (all(is.na(all_posts$month))) all_posts$month = all_posts$window

sum(is.na(all_posts$month))

months.seen = sort(unique(all_posts$month))
counts = table(factor(all_posts$month, levels = months.seen))

plot(seq_along(counts), as.numeric(counts), type = "b",
     xaxt = "n", xlab = "", ylab = "Posts collected",
     main = "Posts collected per month")
axis(1, at = seq_along(counts), labels = names(counts), las = 2, cex.axis = 0.7)
```

Volume here reflects the collection limit as much as the conversation, so any month sitting
at the cap cannot be read as a real total. The share plot below is the honest version.

## Theme Share per Month

```{r trend-share}
tab = table(factor(all_posts$month, levels = months.seen), theme)
share = prop.table(tab, 1)

matplot(share, type = "b", pch = 1, lty = 1:ncol(share), col = 1:ncol(share),
        xaxt = "n", xlab = "", ylab = "Share of posts",
        main = "Theme share of the AI conversation")
axis(1, at = seq_len(nrow(share)), labels = rownames(share), las = 2, cex.axis = 0.7)
legend("topleft", colnames(share), col = 1:ncol(share), lty = 1:ncol(share),
       pch = 1, cex = 0.7, bty = "n")

round(share, 3)
```

# Phase 5: Hypothesis Testing on Engagement

Do posts in one theme attract a different number of likes than posts in another? The two
largest themes are compared.

```{r hypothesis-setup}
theme_sizes = sort(table(theme), decreasing = TRUE)
theme_a = names(theme_sizes)[1]
theme_b = names(theme_sizes)[2]

likes_a = all_posts$like_count[theme == theme_a]
likes_b = all_posts$like_count[theme == theme_b]

cat("Theme A:", theme_a, "-", length(likes_a), "posts\n")
cat("Theme B:", theme_b, "-", length(likes_b), "posts\n")
```

H0: the mean likes for theme A equals the mean likes for theme B.
H1: the means differ. Significance level 0.05.

## Diagnostic Plots

```{r hypothesis-diagnostics, fig.height=4}
par(mfrow = c(1, 2))
qqnorm(likes_a, main = paste("Q-Q plot:", theme_a))
qqline(likes_a, col = "red")
qqnorm(likes_b, main = paste("Q-Q plot:", theme_b))
qqline(likes_b, col = "red")
par(mfrow = c(1, 1))

boxplot(likes_a, likes_b, names = c(theme_a, theme_b),
        ylab = "Number of likes", main = "Likes by theme",
        log = "y")
```

Like counts are heavily skewed, so the Q-Q plots will not be straight. The log scale on the
box plot makes the bulk of the distribution readable.

## Two-Sample Tests

```{r hypothesis-tests}
# Welch t-test on the means. Valid here through the central limit theorem,
# since both samples are large, not because the data is normal.
t_result = t.test(likes_a, likes_b, alternative = "two.sided", var.equal = FALSE)

# Wilcoxon rank-sum compares the distributions without assuming normality.
w_result = wilcox.test(likes_a, likes_b)

cat("Mean likes", theme_a, ":", round(mean(likes_a), 3), "\n")
cat("Mean likes", theme_b, ":", round(mean(likes_b), 3), "\n")
cat("Median likes", theme_a, ":", median(likes_a), "\n")
cat("Median likes", theme_b, ":", median(likes_b), "\n")
cat("t-test p-value:", signif(t_result$p.value, 4), "\n")
cat("Wilcoxon p-value:", signif(w_result$p.value, 4), "\n")
```

Report both. If they disagree, the skew is doing the work, and the Wilcoxon result is the one
to trust. State the conclusion in one sentence: reject H0, or fail to reject H0, at the 0.05
level.

# Phase 6: Analysing Users Rather Than Posts

Clustering users instead of posts allows the text to be compared with how people behave on
the platform. Each user becomes one document made of all their collected posts.

```{r user-documents}
MAX_USERS = 1000

author_counts = sort(table(all_posts$author_handle), decreasing = TRUE)
active_authors = names(author_counts)[1:min(MAX_USERS, length(author_counts))]

# split() groups the text once, which is much faster than filtering per author.
text_by_author = split(all_posts$text, all_posts$author_handle)
user_text = sapply(text_by_author[active_authors], paste, collapse = " ")

length(user_text)
summary(as.numeric(author_counts[active_authors]))
```

## Collecting User Metadata

One API call per user, in batches, with a retry for gateway errors. Set `eval=TRUE` to run,
then leave it off.

```{r user-metadata, eval=FALSE}
user_meta = data.frame(handle = character(), followers_count = numeric(),
                       follows_count = numeric(), posts_count = numeric())

for (i in seq_along(active_authors)) {
  h = active_authors[i]

  info = NULL
  for (attempt in 1:3) {
    info = try(get_user_info(h), silent = TRUE)
    if (!inherits(info, "try-error")) break
    Sys.sleep(attempt * 2)   # back off on a gateway error
  }

  if (inherits(info, "try-error")) {
    user_meta = rbind(user_meta, data.frame(handle = h, followers_count = NA,
                                            follows_count = NA, posts_count = NA))
  } else {
    user_meta = rbind(user_meta, data.frame(handle = h,
                                            followers_count = info$followers_count,
                                            follows_count = info$follows_count,
                                            posts_count = info$posts_count))
  }

  # progress, and a short pause to stay inside the rate limit
  if (i %% 100 == 0) message("users done: ", i, " of ", length(active_authors))
  Sys.sleep(0.2)
}

saveRDS(user_meta, "User_Meta_Data.rds")
```

```{r load-metadata}
user_meta = readRDS("User_Meta_Data.rds")
user_meta = user_meta[complete.cases(user_meta), ]
nrow(user_meta)
```

## Clustering Users by Behaviour

```{r behaviour-clusters}
features = scale(user_meta[, c("followers_count", "follows_count", "posts_count")])

k_values = 2:10
wss = numeric(length(k_values))
sil = numeric(length(k_values))
d_features = dist(features)

for (i in seq_along(k_values)) {
  km = kmeans(features, centers = k_values[i], nstart = 10)
  wss[i] = km$tot.withinss
  sil[i] = mean(silhouette(km$cluster, d_features)[, 3])
}

data.frame(k = k_values, WSS = round(wss, 1), silhouette = round(sil, 3))
```

```{r behaviour-final}
# Set from the table above.
BEHAVIOUR_K = 4

km_behaviour = kmeans(features, centers = BEHAVIOUR_K, nstart = 10)
user_meta$behaviour_cluster = factor(km_behaviour$cluster)

# PCA gives a two-dimensional view of the three behaviour measures.
pca_behaviour = prcomp(features, center = FALSE, scale. = FALSE)
plot(pca_behaviour$x[, 1:2], col = km_behaviour$cluster, pch = 19,
     xlab = "PC1", ylab = "PC2", main = "User behaviour clusters")

# What each cluster looks like.
aggregate(cbind(followers_count, follows_count, posts_count) ~ behaviour_cluster,
          data = user_meta, FUN = median)
table(user_meta$behaviour_cluster)
```

## Clustering Users by Text

```{r user-text-clusters}
user.corpus = Corpus(VectorSource(user_text[user_meta$handle]))
user.corpus = tm_map(user.corpus, function(x) iconv(x, to = "ASCII", sub = " "))
user.corpus = tm_map(user.corpus, remove_urls)
user.corpus = tm_map(user.corpus, removeNumbers)
user.corpus = tm_map(user.corpus, removePunctuation)
user.corpus = tm_map(user.corpus, stripWhitespace)
user.corpus = tm_map(user.corpus, tolower)
user.corpus = tm_map(user.corpus, removeWords, c(stopwords("english"), custom_words))
user.corpus = tm_map(user.corpus, stemDocument)

# Keep terms used by at least five users.
user.dtm = DocumentTermMatrix(user.corpus,
                              control = list(bounds = list(global = c(5, Inf))))
user.tfidf = as.matrix(weightTfIdf(user.dtm, normalize = TRUE))

# Drop users left with no terms.
empties = which(rowSums(abs(user.tfidf)) == 0)
keep_users = user_meta$handle
if (length(empties) > 0) {
  user.tfidf = user.tfidf[-empties, ]
  keep_users = keep_users[-empties]
}

dim(user.tfidf)
```

```{r user-text-k}
user.norm = user.tfidf / sqrt(rowSums(user.tfidf^2))
D_user = dist(user.norm, method = "euclidean")^2 / 2

wss_u = numeric(length(k_values))
sil_u = numeric(length(k_values))

for (i in seq_along(k_values)) {
  km = kmeans(user.norm, centers = k_values[i], nstart = 10)
  wss_u[i] = km$tot.withinss
  sil_u[i] = mean(silhouette(km$cluster, D_user)[, 3])
}

data.frame(k = k_values, WSS = round(wss_u, 2), silhouette = round(sil_u, 3))
```

```{r user-text-final}
# Set from the table above.
TEXT_K = 3

km_text = kmeans(user.norm, centers = TEXT_K, nstart = 25)
table(km_text$cluster)

# Terms that describe each user cluster.
for (cl in 1:TEXT_K) {
  weights = colMeans(user.tfidf[km_text$cluster == cl, , drop = FALSE])
  o = order(weights, decreasing = TRUE)[1:10]
  cat("User cluster", cl, "(", sum(km_text$cluster == cl), "users):",
      colnames(user.tfidf)[o], "\n")
}
```

## Comparing the Two Clusterings

```{r cluster-comparison}
behaviour_of_kept = user_meta$behaviour_cluster[match(keep_users, user_meta$handle)]

contingency = table(text_cluster = km_text$cluster, behaviour_cluster = behaviour_of_kept)
contingency

chisq.test(contingency)
```

A small p-value means what users write is associated with how they behave on the platform. A
large one means the two views of the same people are telling different stories, which is also
a finding worth reporting.

# Phase 7: Network and Link Analysis

## Choosing the Seed User

```{r seed-user}
seed_user = user_meta$handle[which.max(user_meta$followers_count)]
cat("Seed user:", seed_user, "with", max(user_meta$followers_count), "followers\n")
```

## Collecting Follows

```{r follows, eval=FALSE}
FIRST_DEGREE_LIMIT = 25

first_degree = get_follows(seed_user)
first_degree = head(first_degree, FIRST_DEGREE_LIMIT)

edges = data.frame(from = seed_user, to = first_degree$actor_handle,
                   stringsAsFactors = FALSE)

for (h in first_degree$actor_handle) {
  friends = try(get_follows(h), silent = TRUE)
  if (!inherits(friends, "try-error") && nrow(friends) > 0) {
    edges = rbind(edges, data.frame(from = h, to = friends$actor_handle,
                                    stringsAsFactors = FALSE))
  }
  Sys.sleep(0.5)
}

saveRDS(edges, "Follow_Edges.rds")
```

```{r build-graph, fig.height=6}
library(igraph)

edges = readRDS("Follow_Edges.rds")

g = graph_from_edgelist(as.matrix(edges), directed = TRUE)
g = simplify(g)

vcount(g)
ecount(g)
edge_density(g)

# Only the better connected part is readable in a plot.
g_core = induced_subgraph(g, which(degree(g, mode = "all") > 2))

plot(g_core, layout = layout_with_fr(g_core), vertex.size = 5,
     vertex.label = NA, edge.arrow.size = 0.2, vertex.color = "steelblue",
     main = "Follow network around the seed user")
```

## Centrality and PageRank

```{r centrality}
deg = degree(g, mode = "all")
bet = betweenness(g, directed = TRUE)
pr = page_rank(g)$vector

central = data.frame(handle = names(deg),
                     degree = deg,
                     betweenness = round(bet, 1),
                     pagerank = round(pr, 5))

head(central[order(central$degree, decreasing = TRUE), ], 10)
head(central[order(central$betweenness, decreasing = TRUE), ], 10)
head(central[order(central$pagerank, decreasing = TRUE), ], 10)
```

High degree with high betweenness marks an account that is both popular and a bridge between
groups. High betweenness with low degree marks a gatekeeper, an account that few people
follow but which connects otherwise separate parts of the network.

# What to Decide Next

- Choose one or two themes that hold enough posts and show movement across the months.
- Run a focused collection on those themes with narrower phrases, using the same monthly loop.
- Add the network work from Modules 7 and 8 on the chosen theme: seed account, follows,
  centrality and PageRank.

# Limitations

- Each search is capped per month and per phrase, so counts measure collection effort as much
  as conversation volume. Read shares, not totals.
- Within a capped month, `sort = "latest"` returns posts from the end of that month, so the
  sample inside a capped window is not random.
- Only English posts are kept, and Bluesky users are not a cross-section of the public.
- Stemming merges word forms, so themes and cluster terms read as fragments.
- Themes are built from the top terms of a 2,000 post sample, then applied to every post.
- Theme labelling is first match wins, so overlapping posts land in the earlier theme. Report
  how many posts fall into "other".
